Long name of the paper

### 2018-10-17 anzjs2018/sections/02_gtfs.tex
edits

GTFS (general transit feed specification)
is an API (application programming interface) specification for transit data,
developed and maintained by Google \citep{GoogleDevelopers_2006}.
It is used by over 900~transit providers around the world,
including here in Auckland, New Zealand
(source \url{http://transitfeeds.com}),
allowing an application developed in Auckland to
be deployed to another GTFS-based public transport system with minimal modification.

The two main components of GTFS are \emph{static} and \emph{realtime}.
Static GTFS includes information about routes, trips, shapes, and stops:
a \emph{route} is a sequence of two or more stops displayed as a single service;
a \emph{trip} is an instance of a route occuring at a specific time of day;
a \emph{shape} is the sequence of points defining a vehicle's path along a route;
and a \emph{stop} is a physical location where passengers can embark and disembark
the vehicle \citep{GoogleDevelopers_2006}.
GTFS realtime feeds provide information such as vehicle positions and trip delays,
and are usually accessed via an API.
% The latter is often the only covariate used for predicting arrival time,
% as previously discussed:
% while it is a computationally inexpensive method (simply addition),
% it relies on accurate schedules and cannot respond quickly to realtime events.

@@

Arguably one of the most important predictors of arrival time is
the travel time along intermediate roads,
however in most applications this vital information is unavailable,
at least directly. 
Several predictive approaches used \emph{headway},
the time between consecutive trips along the same route
\citep{Hans_2015}.
While this is reasonable for high frequency routes,
low frequncy routes will be unable to react quickly to changes in congestion,
and it is for these routes which reliable ETAs are arguably more important,
since the cost of missing a bus is greater.

One solution would be to use information obtained from
vehicles servicing other routes but traveling along the same roads
to estimate arrival times;
however, there is no direct way of knowing which routes share common roads.
Here we propose using an algorithm to convert the raw GTFS shape files
into a \emph{road network},
as demonstated in Figure~\ref{fig:network_creation}.
This involves detecting locations where one or more routes meet 
(intersections or \emph{nodes}),
and the connecting paths (road segments or \emph{edges}).
However, in this paper we use bus stops as nodes and the routes between them as edges.
This captures most overlaps, 
with the exception of some express services and multiple-stop locations,
but these are typically in high traffic areas and should be OK.
Figure~\ref{fig:network_creation_2} demonstrates how several overlapping routes 
are merged to form a road network.
In section~\ref{sec:kf} we present a model for estimating veihcle travel time
along a road segment.

@@

GTFS \rt allows developers to query the current positions of vehicles
in the transit network.
The data consists of the time $t_k$ that the observation was made,
the GPS position of the vehicle, $\bY_k$, 
and some other information about the trip being serviced.
Vehicle positions are updated with a frequency of anywhere from 10~seconds to several minutes,
so there is often a lot of uncertainty about the trajectory
between two observations, particularly when there is a bus stop
or intersection between them.

Another complication with the Auckland Transport realtime feed is that
the buses are programmed to report their location when arriving at
bus stops and some major intersections.
Often these positions appear to be preemptive (i.e., that bus is almost there, 
but not quite),
and subsequent observations place the bus \emph{behind} the stop or intersection.
To handle this, we compute the approximate distance traveled, $\tilde x_k$,
of the vehicle by finding the nearest point on the path to the observation;
if this has decreased, the previous observation is rejected and the vehicle reverted
to its previous state.
To enable this, vehicles retain the current and previous states (see section~\ref{sec:rt}).

>>>>

GTFS (general transit feed specification)
is an API (application programming interface) specification for transit data
detailing it should be organised,
making access easier for application developers.
Developed and maintained by Google \citep{GoogleDevelopers_2006},
who make use of it in their Google Maps Transit Directions,
it is used by over 900~transit providers around the world,
including here in Auckland, New Zealand
(source \url{http://transitfeeds.com}).
An advantage of this for us is that,
provided our application depends solely on GTFS data,
after developing it locally in Auckland it can be deployed to any other GTFS-based
public transport system with minimal modification.

There are two components to GTFS.
The first, \emph{GTFS static}, includes information about stops, routes, trips, and shapes:
a \emph{stop} is a physical location where passengers can embark and disembark the vehicle;
a \emph{route} is a sequence of two or more stops displayed as a single service;
a \emph{trip} is an instance of a route occuring at a specific time of day; and
a \emph{shape} is the sequence of points defining a vehicle's path along a route
\citep{GoogleDevelopers_2006}.
The second component is \emph{GTFS realtime},
which is only available in a subset of the providers due to the requirement of 
onboard GPS and a central server.
It provides a standardised format for sharing vehicle positions and trip delays,
which are often accessed via an API which developers can use in \rt applications.

Before we can develop our application,
we need to modify the raw GTFS information to make it possible to detect common roads.

@@

Arguably one of the most important predictors of arrival time is
the travel time along intermediate roads,
however in most applications this vital information is unavailable,
at least directly. 
Several predictive approaches using \emph{headway},
the time between consecutive trips along the same route,
have been proposed \citep{Hans_2015};
however, while this is reasonable for high frequency routes,
low frequncy routes will be unable to react quickly to changes in congestion,
and it is for these routes which reliable ETAs are arguably more important,
since the cost of missing a bus is greater.

One solution would be to use information obtained from
vehicles servicing other routes but traveling along the same roads
to estimate arrival times.
The simplest method of identifying overlapping routes is to compare
the stop sequence: 
routes with a common sub-sequence of stops must be traveling the same route between them.
While there are several exceptions, for example express routes 
and multi-stop locations where there are high numbers of routes,
it is a viable enough simplication for the purposes of this paper.
Figure~\ref{fig:network_creation} demonstrates how several overlapping routes 
are merged to form a road network.
In section~\ref{sec:kf} we present a model for estimating vehicle travel time
along a road segment.

@@

GTFS \rt allows developers to query the current positions of vehicles
in the transit network.
The data consists of the time $t_k$ that the observation was made,
the GPS position of the vehicle, $\bY_k$, 
and some other information about the trip being serviced.
Vehicle positions are updated with a frequency of anywhere between 10~seconds and several minutes,
so there is often a lot of uncertainty about the trajectory
between two observations, particularly when there is one or more bus stop
or intersection between them.

Another complication with the Auckland Transport realtime feed is that
the buses are programmed to report their location when arriving at
bus stops and some major intersections.
Often these positions appear to be preemptive 
(i.e., the bus is almost there, but not quite),
and subsequent observations place the bus \emph{behind} the stop or intersection
(e.g., in a queue of traffic at traffic lights).
To handle this, we compute the approximate distance traveled, $\tilde x_k$,
of the vehicle by finding the nearest point on the path to the observation;
if this has decreased, the previous observation is rejected and the vehicle reverted
to its previous state.

### 2018-10-17 anzjs2018/sections/03_models.tex
edits

The real-time nature of this application suggests that
a recursive Bayesian filtering approach would be appropriate, as is often used in
real-time tracking and robotics applications.
We are using two separate models, one each for the vehicle and network states.
In each,
we assume an underlying Markov process with state $\mathcal{X}_k$,
of which observations $\mathcal{Y}_k$ are made,
giving us to the following general model,
\begin{equation}
\label{eq:rbe_model}
\begin{split}
\mathcal{X}_k &= f(\mathcal{X}_{k-1}, \omega_k) \\
\mathcal{Y}_k &= h(\mathcal{X}_k, \nu_k)
\end{split}
\end{equation}
where $\omega_k\sim\mathrm{N}(0,\mathcal{Q})$ is the system noise,
and $\nu_k\sim\mathrm{N}(0,\mathcal{R})$ is measurement error.
The transition function $f$ determines the relationship between consecutive states,
while the measurement function $h$ is a deterministic function describing the
relationship between the underlying and observed states.

The following sections describe the two Bayesian filter models based on (\ref{eq:rbe_model}) used in this application.
The first is implemented using a particle filter to estimate vehicle states,
with the primary objective of estimating vehicle travel times along roads,
while the second using these estimated times to update road states,
and is implemented using a variant of the Kalman filter referred to as the \emph{information filter}
\citep{}.

@@

The underlying vehicle state at time $t_k$ consists of
the vehicle's distance traveled $x$ in meters along the route and
its speed $\dot x$ in meters per second.
Additionally, we include travel times along road segments along the route, 
$\bz = (z_1,\ldots,z_L)^\top$, in seconds,
which are independent of time but included in the state as they are estimated
sequentially as the vehicle traverses the route.
These are combined into the state vector,
\begin{equation}
\label{eq:vehicle_state}
\bX_k = 
\begin{bmatrix}
    x_k \\ \dot x_k \\ \bz
\end{bmatrix}.
\end{equation}
Observations of the vehicle are made using a GPS at time $t_k$,
giving the longitude $\lambda_k$ and latitude $\phi_k$ of the vehicle
and stored in the observation vector
\begin{equation}
\label{eq:vehicle_obs}
\bY_k = \begin{bmatrix} \lambda_k \\ \phi_k \end{bmatrix}.
\end{equation}

We chose to employ a particle filter to estimate $\bX_k$,
to do its high flexibility and use in recent 
transit vehicle modeling applications \citep{Hans_2015}.
Our main justification for using it is that it handles multimodality very well,
as demonstrated in figure~\ref{fig:pf_state},
which is a common feature of the proposal distribution particularly around bus stops.
Another advantage is the intuitive likelihood function (see section~\ref{sec:pf_update}).
Conversely, the particle filter is a computationally demanding method,
as thousands of particles are required per bus.
Section~\ref{sec:rt} describes how we were able to implement the particle filter in real-time and the timings.

In the particle filter, the posterior distribution of the state at time $t_{k-1}$,
is represented by a set of discrete points, or particles, (fig~\ref{fig:pf_state_prev})
\begin{equation}
p(\bX_{k-1} | \bY_{k-1}) \approx \tilde\bX_{k-1|k-1} := (\bX_{k-1}^{(i)})_{i=1}^N
\end{equation}
each of which is independently updated or \emph{mutated} using the transition function $f$ (fig~\ref{fig:pf_state_predict})
\begin{equation}
p(\bX_k | \bX_{k-1}) \approx \tilde\bX_{k|k-1} := 
\left(f(\bX_{k-1}^{(i)}, \psi)\right)_{i=1}^N
\end{equation}
using a parameter vector $\psi$ containing all of the necessary parameters
for the model (including system noise).
After mutation, \emph{selection} of a new set of particles is performed by
importance resampling using likelihood weights,
determined by distance between particle's map position and $\bY_k$ 
to obtain a new sample $\tilde\bX_{k|k}$, (fig~\ref{fig:pf_state_update}).

@@

@@

Another advantage of using a particle filter to implement the vehicle model is that
the transition function $f$ can implement several bus behaviours,
most importantly
\begin{itemize}
\item non-constant speed along roads (acceleration as system noise),
\item optional stopping and waiting at bus stops while passengers board and disembark, and
\item optional stopping and waiting at intersections.
\end{itemize}

For each particle, the transition function generates a plausible trajectory,
using system noise parameter $\sigma^2$ which describes 
how the vehicles acceleration changes as a random process.
However, the vehicle does not travel constantly along the route,
as it needs to service bus stops along the way.
Therefore, the transition function includes 
stopping probabilities $\boldsymbol\pi = (\pi_1,\ldots,\pi_J)^\top$ at bus stops,
dwell times $\boldsymbol\tau = (\tau_1,\ldots,\tau_J)^\top$ for passengers to
board and disembark (conditional on the bus stopping),
and the minimum dwell time at stops, $\gamma$.
For these parameters, we used constant values for all stops,
and based the values on those used by \cite{Hans_2015};
future work will look at modeling these separately in real-time similarly to \ref{sec:kf}.
As we are using bus stops to determined segments,
no intersection model is required,
only the segment travel times as predicted by the network model at time $t_k$,
$\btheta(t_k) = (\theta_1(t_k), \ldots, \theta_L(t_k))^\top$
(see section~\ref{sec:kf}).
Future work will adapt this to use intersection that are indpendent 
of bus stops, and therefore completely independent of routes,
which will allow increasing the complexity of the model to incorporate
intersection stopping probabilities and waiting times.
The details of the transition function are given in the algorithm defined in the Appendix,
which implements those features discussed above.

@@

After mutating the particle set, the posterior distribution $\tilde\bX_{k|k}$ is obtained by
\emph{selection}, which is implemented by importance resampling using
likelihood weights.
To do so, we define the measurement function $h$,
and an additional function $g$ which transforms GPS coordinates onto a flat
surface (the geographical equirectangular projection),
which allows us to place a bivate normal likelihood on the data
as shown in figure~\ref{fig:pf_lhood},
The model for the observation generation is,
assuming GPS error $\epsilon^2$,
\begin{equation}
\label{eq:pf_obs_model}
g(\bY_k) = g(h(\bX_k)) + \br_k,
\quad \br_k \sim \mathrm{N}(\boldsymbol{0}, \epsilon^2\boldsymbol{I})
\end{equation}
However, due to the assumption that error in longitude and latitude are uncorrelated, 
the geographical distance between the observation and the true vehicle can be expressed
in terms of two independent standard normal random variables $z_1$ and $z_2$
\begin{equation}
\label{eq:obs_dist}
dist(\bY_k, h(\bX_k)) = \sqrt{r_{k1}^2 + r_{k2}^2} 
    = \sqrt{(\epsilon z_1)^2 + (\epsilon z_2)^2}
\end{equation}
As the sum of two independent, standard normal random variables 
is $\chi^2$ distributed with 2~degrees of freedom,
which is itself exponential,
rearranging (\ref{eq:obs_dist}) yields
\begin{equation}
\label{eq:obs_exp}
\left(\frac{dist(\bY_k, h(\bX_k))}{\epsilon}\right)^2 =
z_1^2 + z_2^2 \sim \mathrm{Exp}\left(\frac{1}{2}\right)
\end{equation}

The likelihood of the data given a particle's state estimate 
is therefore easy to calculate using (\ref{eq:obs_exp})
\begin{equation}
p(\bY_k | \bX_k^{(i)}, \epsilon) =
\frac{1}{2}\exp\left\{
-\frac{1}{2} \left(\frac{dist(\bY_k, h(\bX_k^{(i)}))}{\epsilon}\right)^2
\right\}
\end{equation}
It is worth noting that this representation of the likelihood is only
possible within the particle filter;
in other approaches, such as the Kalman filter,
the likelihood cannot be calculated using the distance as the estimate
is a distribution, and so instead a reverse non-deterministic transformation
is required, which introduces additional error and uncertainty into the model.

The posterior distribution is obtained by \emph{selection},
or resampling with replacement using the likelihood weights
\begin{equation*}
W_k^{(i)} = \frac{p(\bY_k | \bX_k^{(i)}, \epsilon)}{
    \sum_{j=1}^N p(\bY_k | \bX_k^{(j)}, \epsilon)
} 
W_{k-1}^{(i)} 
\end{equation*}
However, to improve efficiency, the state $\tilde \bX_k$ is only resampled
when the effective sample size 
$N_{\text{eff}}$ is less than some threshold,
\begin{equation*}
N_{\text{eff}} = \frac{1}{\sum_{i=1}^N (W_k^{(i)})} < N_{\text{thres}}
\end{equation*}
When resampling is performed, $W_k^{(i)} := N^{-1}$ for all $i$.
When sampling is not performed,
state estimates use $\{\bX_k^{(i)}, W_k^{(i)}\}$ to obtain weighted estimates.

@@

[[  blurb  ]]
% The second part of this appraoch involves a network model of transit network travel times along links in the network.
% Reliable ETAs will require a historical data component,
% used as a prior in the absense of data as well as for
% predicting future states, 
% and a realtime update component where the travel times of vehicles
% are used to update the current estimate of network state.

To obtain reliable ETAs,
it is necessary to estimate travel times along roads,
as this allows the model to react to \rt congestion as detected by other buses.
Future work will aim to construct a more complex model incorporating historical trends 
and correlations between roads, 
however for the current work we are demonstrating our approach
using a simple model.

@@

The network state $\boldsymbol\theta_c = \{\theta_c^\ell\}_{\ell = 1}^L$ is the travel time 
of transit vehicles along road segment $\ell$ at time $t_c$.
Observations of the state, $Z_c^\ell$, are obtained from the particle filter above.
We are assuming independence between each segment
to provide a simple basis for the model; of course in reality this
is not true and future work will investigate ways of incorporating this.
Thanks to the model simplicity,
no transition matrix or measurement matrix is required,
and the model from (\ref{eq:rbe_model}) reduces to
\begin{equation}
\begin{split}
\hat \theta_c^\ell &= \theta_{c-1}^\ell + \Delta_c^\ell v_c^\ell \\
Z_c^\ell &= \theta_c^\ell + e_c^\ell
\end{split}
\end{equation}
where $\Delta_c = t_c - t_{c-1}$,
$v_c^\ell \sim \mathrm{N}(0, \nu^2)$ is the system noise,
and $e_c^\ell \sim \mathrm{N}(0, R_c^\ell)$ is the measurement error
as estimated from the vehicle model using particle uncertainty.

@@

For the update step, the parameters are transformed into an information
space parameterised by the information matrix $U^\ell_c$
and the information vector $u^\ell_c$,
\begin{equation}
\label{eq:information_transform}
U^\ell_{c|c-1} = \frac{1}{P_{c|c-1}}
\quad\text{and}\quad
\hat u^\ell_{c|c-1} = \frac{\hat \theta^\ell_{c|c-1}}{P^\ell_{c|c-1}}
\end{equation}
The measurement data obtained from the vehicle model are
the travel time of vehicle $m$ along segment $\ell$,
$\bar z_c^{m\ell}$, and the uncertainty $s^{m\ell}_c$.
These are transformed to a measurement information covariance matrix $B$
and vector $b$ using
\begin{equation*}
B^{m\ell}_c = \frac{1}{(s^{m\ell}_c)^{2}}\quad\text{and}\quad
b^{m\ell}_c = \frac{\bar z^{m\ell}_c}{(s^{m\ell}_c)^2}
\end{equation*}
The information update is now the sum of the information for all vehicles
that traversed the segment since the last update, so
\begin{align*}
U^\ell_{c|c} &= U^\ell_{c|c-1} + \sum_{m=1}^M B^{m\ell}_{c} \\
\hat u^\ell_{c|c} &= \hat u^\ell_{c|c-1} + \sum_{m=1}^M b^{m\ell}_{c}
\end{align*}
which are back transformed to get the updated state parameters
using the inverse of (\ref{eq:information_transform})
\begin{equation*}
\hat \theta^\ell_{c|c} = \frac{\hat u^\ell_{c|c}}{U^\ell_{c|c}} 
\quad\text{and}\quad
P^\ell_{c|c} = \frac{1}{U^\ell_{c|c}}.
\end{equation*}

>>>>

Due to the real-time nature of this application, we are using
a recursive Bayesian filtering approach as is often used in
real-time tracking applications.
We are using two distinct models for the vehicle and network states,
and in each assume an underlying Markov process with state $\mathcal{X}_k$
with system noise $\omega_k\sim\mathrm{N}(0,\mathcal{Q})$ 
of which observations $\mathcal{Y}_k$ are made
with error $\nu_k\sim\mathrm{N}(0,\mathcal{R})$,
giving the following general model below.
\begin{equation}
\label{eq:rbe_model}
\begin{split}
\mathcal{X}_k &= f(\mathcal{X}_{k-1}, \omega_k) \\
\mathcal{Y}_k &= h(\mathcal{X}_k, \nu_k)
\end{split}
\end{equation}
The transition function $f$ determines the relationship between consecutive states,
while the measurement function $h$ is a deterministic function describing the
relationship between the underlying and observed states.

The following sections describe the two models used in this application.
The first is implemented using a particle filter to estimate vehicle states,
with the primary objective of infering vehicle travel times along roads
from a sequence of GPS positions,
while the second uses these travel times to update road states 
and is implemented using an \emph{information filter},
which is a variant of the Kalman filter \citep{}.

@@

The underlying vehicle state at time $t_k$ consists of
the vehicle's distance traveled $x$ in meters along the route and
it's speed $\dot x$ in meters per second.
Additionally, travel times along road segments along the route, 
$\bz = (z_1,\ldots,z_L)^\top$, in seconds,
are independent of time but included in the state so they can be estimated
sequentially as the vehicle traverses the route.
These are combined into the state vector,
\begin{equation}
\label{eq:vehicle_state}
\bX_k = 
\begin{bmatrix}
    x_k \\ \dot x_k \\ \bz
\end{bmatrix}.
\end{equation}
Observations $\bY_k$ of the vehicle are made using a GPS at time $t_k$
with GPS error $\epsilon^2$,
giving the longitude $\lambda_k$ and latitude $\phi_k$ of the vehicle,
\begin{equation}
\label{eq:vehicle_obs}
\bY_k = \begin{bmatrix} \lambda_k \\ \phi_k \end{bmatrix}.
\end{equation}

We chose to employ a particle filter to estimate $\bX_k$,
due to do its high flexibility and use in recent 
transit vehicle modeling applications \citep{Hans_2015}.
The primary advante of a particle filter is its handling of multimodality,
as in figure~\ref{fig:pf_state_predict},
which is a common feature of the proposal distribution, particularly around bus stops.
A second advantage is the intuitive likelihood function (see section~\ref{sec:pf_update}).
Conversely, particle filter methods are computationally demanding,
requiring an increasing number of particles as the model complexity and
number of parameters in increase \citep{}.
Section~\ref{sec:rt} describes our implementation of the particle filter in real-time 
and it's timings.

@@

In a particle filter, the posterior distribution of the state at time $t_{k-1}$,
is represented by a set of discrete points, or particles, (fig~\ref{fig:pf_state_prev}),
each with an associated weight $W_{k-1}^{(i)}$
\begin{equation}
p(\bX_{k-1} | \bY_{k-1}) \approx \tilde\bX_{k-1|k-1} 
:= \{\bX_{k-1}^{(i)}, W_{k-1}^{(i)}\}_{i=1}^N
\end{equation}
each of which is independently updated or \emph{mutated} using the transition function $f$ (fig~\ref{fig:pf_state_predict})
\begin{equation}
p(\bX_k | \bX_{k-1}) \approx \tilde\bX_{k|k-1} := 
\left\{f(\bX_{k-1}^{(i)}, \psi), W_{k-1}^{(i)}\right\}_{i=1}^N
\end{equation}
using parameter vector $\psi$ containing all of the necessary parameters
for the model (including system noise).
After mutation, 
the particles are reweighted using the likelihood $p(\bY_k | \bX_k^{(i)})$ 
(sec.~\ref{sec:pf_update}) and standardised,
\begin{equation*}
W_k^{(i)} = \frac{W_{k-1}^{(i)} p(\bY_k | \bX_k^{(i)})}{
    \sum_{j=1}^N W_{k-1}^{(j)} p(\bY_k | \bX_k^{(j)})
}
\end{equation*}
generating a posterior estimate of the state
\begin{equation}
p(\bX_k | \bY_k) \approx \tilde\bX_{k|k} := 
\{\bX_{k}^{(i)}, W_{k}^{(i)}\}_{i=1}^N.
\end{equation}

In order to prevvent degeneration of the particle sample,
\emph{importance resampling} is employed whenever the 
effective sample size $N_{\text{eff}} = 1/(\sum_i (W_k^{(i)})^2)$
falls below a specified threshold $N_{\text{thres}}$.
The resampling is performed with replacement using importance weights $W_k^{(i)}$,
as demonstrated in Figure~\ref{fig:pf_state_update}.
Afterwards, the weights are reinitialized to $W_k^{(i)} = N^{-1}$.

@@

Using a particle filter allows us to develop a vehicle model in which
the transition function $f$ implements several key bus behaviours:
\begin{itemize}
\item non-constant speed along roads,
\item optional stopping and waiting at bus stops while passengers board and disembark, and
\item optional stopping and waiting at intersections/traffic lights (future work).
\end{itemize}

For each particle, the transition function generates a plausible trajectory,
using system noise parameter $\sigma^2$ which describes 
how the vehicle's acceleration changes as a random process.
The vehicle does not travel constantly along the route,
however, as it needs to service bus stops along the way.
Therefore, the transition function includes 
stopping probabilities $\boldsymbol\pi = (\pi_1,\ldots,\pi_J)^\top$ at bus stops,
dwell times $\boldsymbol\tau = (\tau_1,\ldots,\tau_J)^\top$ for passengers to
board and disembark (conditional on the bus stopping),
and the minimum dwell time at stops, $\gamma$,
for the doors to open and close.
For these parameters, we used constant values for all stops,
and based the values on those used by \cite{Hans_2015};
future work will investigate modeling these in real-time similarly to Section~\ref{sec:kf}.
No intersection model is implemented here,
as the segment road segments are between stops,
and the travel times predicted by the network model at time $t_k$,
$\btheta(t_k) = (\theta_1(t_k), \ldots, \theta_L(t_k))^\top$
(see section~\ref{sec:kf}).
The transition function algorithm is shown in Figure~\ref{fig:algorithm},
which implements those features discussed above.

@@

After mutating the particle set, the posterior distribution $\tilde\bX_{k|k}$ 
is updated by reweighting each particle using the likelihood.
The measurement function $h$,
and an additional function $g$ which transforms GPS coordinates onto a flat
surface (the geographical equirectangular projection),
allow the placement of a bivate normal likelihood on the data
as shown in figure~\ref{fig:pf_lhood}.
The model for the observation generation is,
assuming GPS error $\epsilon^2$,
\begin{equation}
\label{eq:pf_obs_model}
g(\bY_k) = g(h(\bX_k)) + \br_k,
\quad \br_k \sim \mathrm{N}(\boldsymbol{0}, \epsilon^2\boldsymbol{I})
\end{equation}
Now, the geographical distance between the observation and the true vehicle can be expressed
in terms of two independent random variables $z_1, z_2 \sim \mathrm{N}(0,1)$
\begin{equation}
\label{eq:obs_dist}
dist(\bY_k, h(\bX_k)) = \sqrt{r_{k1}^2 + r_{k2}^2} 
    = \sqrt{(\epsilon z_1)^2 + (\epsilon z_2)^2}.
\end{equation}
The sum of two independent, standard normal random variables 
is $\chi^2$ distributed with 2~degrees of freedom,
which is itself exponential,
so rearranging (\ref{eq:obs_dist}) yields
\begin{equation}
\label{eq:obs_exp}
\left(\frac{dist(\bY_k, h(\bX_k))}{\epsilon}\right)^2 =
z_1^2 + z_2^2 \sim \mathrm{Exp}\left(\frac{1}{2}\right)
\end{equation}

The likelihood of the data given a particle's state estimate 
is therefore easy to calculate using (\ref{eq:obs_exp})
\begin{equation}
p(\bY_k | \bX_k^{(i)}, \epsilon) =
\frac{1}{2}\exp\left\{
-\frac{1}{2} \left(\frac{dist(\bY_k, h(\bX_k^{(i)}))}{\epsilon}\right)^2
\right\}
\end{equation}
It is worth noting that this representation of the likelihood is only
possible due to the discretization of the state provided by the particle filter;
in other approaches, such as the Kalman filter,
the likelihood cannot be calculated using the distance as the estimate
is a distribution, and so instead a reverse non-deterministic transformation,
the inverse measurement function,
is required to estimate the ``observed'' distance traveled $\hat Z_k~=~h^{-1}(\bY_k)$, 
which introduces additional error and uncertainty into the model.

@@

The purpose of the road network model is to provide a way of using
spatial information about vehicle speeds to inform arrival time predictions.
The simplest notion would be to map each vehicle speed estimate to a physical road;
however, this has two problems
\begin{enumerate}
\item roads usually have two directions, so vehicle directionality would need to be computed,
\item road data is not provided in GTFS, and can be difficult to obtain cleanly without
    manual intervention.
\end{enumerate}
The most desirable solution would be to use the GTFS shape data to construct a network
by merging overlapping routes and using merge points as intersections
to define road segments.
Initial efforts at applying this had poor results due to overly long or short segments,
so for this current work we are approximating the road network by using the stops 
along a route as intersection, and the legs between them as road segments,
as shown in Figure~\ref{fig:network_creation}.

To obtain reliable ETAs,
road travel times need to be estimated in \rt,
as this allows the model to react to congestion changes as detected by preceeding buses.
Future work will aim to construct a more complex model incorporating historical trends 
and correlations between roads, 
however for the current work we are demonstrating our approach
using a simple model,
in which each road segment is treated independently.

@@

The network state $\boldsymbol\theta_c = \{\theta_c^\ell\}_{\ell = 1}^L$ is the travel time 
of transit vehicles along road segment $\ell$ at time $t_c$.
Observations of the state, $Z_c^\ell$, are obtained from the particle filter 
in Section~\ref{sec:pf}.
Due to the simplicity of the model,
no transition matrix or measurement matrix is required,
and the model from (\ref{eq:rbe_model}) reduces simply to
\begin{equation}
\begin{split}
\theta_c^\ell &= \theta_{c-1}^\ell + \Delta_c^\ell v_c^\ell \\
Z_c^\ell &= \theta_c^\ell + e_c^\ell
\end{split}
\end{equation}
where $\Delta_c = t_c - t_{c-1}$,
$v_c^\ell \sim \mathrm{N}(0, \nu^2)$ is the system noise,
and $e_c^\ell \sim \mathrm{N}(0, R_c^\ell)$ is the measurement error
as estimated from the vehicle model using uncertainty of particle estimates.

@@

For the update step, the parameters are transformed into an information
space parameterised by the information matrix $U^\ell_c$
and the information vector $u^\ell_c$,
\begin{equation}
\label{eq:information_transform}
U^\ell_{c|c-1} = \frac{1}{P_{c|c-1}}
\quad\text{and}\quad
\hat u^\ell_{c|c-1} = \frac{\hat \theta^\ell_{c|c-1}}{P^\ell_{c|c-1}}
\end{equation}
The measurement data obtained from the vehicle model are
the travel time of vehicle $m$ along segment $\ell$,
$\bar z_c^{m\ell}$, and the uncertainty $s^{m\ell}_c$.
These are transformed to a measurement information covariance matrix $B_c^{m\ell}$
and measurement information vector $b_c^{m\ell}$,
\begin{equation*}
B^{m\ell}_c = \frac{1}{(s^{m\ell}_c)^{2}}\quad\text{and}\quad
b^{m\ell}_c = \frac{\bar z^{m\ell}_c}{(s^{m\ell}_c)^2}
\end{equation*}
The information update is now the sum of the information over all $M$ vehicles
that traversed the segment since the last update, so
\begin{align*}
U^\ell_{c|c} &= U^\ell_{c|c-1} + \sum_{m=1}^M B^{m\ell}_{c} \\
\hat u^\ell_{c|c} &= \hat u^\ell_{c|c-1} + \sum_{m=1}^M b^{m\ell}_{c}
\end{align*}
which are back transformed
using the inverse of (\ref{eq:information_transform})
to obtain the desired parameter estimates
\begin{equation*}
\hat \theta^\ell_{c|c} = \frac{\hat u^\ell_{c|c}}{U^\ell_{c|c}} 
\quad\text{and}\quad
P^\ell_{c|c} = \frac{1}{U^\ell_{c|c}}.
\end{equation*}
